---
title: "CT Housing Data"
author: Magdalena Kossuth
format: html
editor: visual
---

```{r}
#| warning: false
#| echo: false

library(dplyr)
library(scales)
library(stringr)
library(tidyverse)
library(visdat)
theme_set(theme_bw())
```

From the Open CT Data Website: Monthly sale price of single-family homes in Connecticut, 2001 through the present. Data updated monthly by the Connecticut Housing Finance Authority and tracked in the following dashboard: <https://www.chfa.org/about-us/ct-monthly-housing-market-dashboard/>.

```{r}
#| warning: false
#| echo: false
#| cache: true
#| label: load data

housing_data <- read_csv("https://data.ct.gov/resource/fv7n-cnqr.csv?$query=SELECT%20%60date%60%2C%20%60median_sale_price%60%2C%20%60average_sale_price%60") 

glimpse(housing_data)
```

\
How have the average housing prices changed over the years?

```{r}
#| echo: false
#| label: houses prices bar graph
 
housing_yearly <- housing_data |>
  mutate(year = year(date)) |>
  group_by(year) |>
  summarise(
    median_price = median(median_sale_price, na.rm = TRUE),
    average_price = mean(average_sale_price, na.rm = TRUE))

ggplot(housing_yearly, aes(x = as.factor(year), y = average_price)) +
  geom_col(color = "darkblue", fill = "cornflowerblue") +
  labs(title = "Average Home Prices in CT from Years 2006 to 2022",
       y = "Average Price", x = "Year") +
  scale_y_continuous(labels = label_number(scale = 1e-3, suffix = "K")) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

The bar graph exhibits a parabolic pattern.

At the beginning of 2019, when COVID-19 reached the US, housing sales and new construction initially halted. However, since then the housing market in CT became more and more competitive. Is this increase in house prices significant?

To answer this question let's conduct a one-way ANOVA test.

The null hypothesis: the means do not differ.

Alternative hypothesis: the means are different from one another.

```{r}
#| label: anova
#| eval: false

anova <- aov(average_price ~ year, data = housing_yearly)
summary(anova)
```