---
title: "Thyroid Cancer Recurrence"
format: html
editor: visual
---

This dataset focuses on thyroid cancer recurrence after Radioactive Iodine (RAI) therapy. It contains 383 patient records with 13 key attributes. This data will be used to build a valuable predictive model for cancer recurrence.

\
**Column Descriptions:**

-   **Age** : Age of the patient (in years).

-   **Gender** : Patient's gender (Male or Female).

-   **Hx Radiotherapy** : History of prior radiotherapy (Yes or No).

-   **Adenopathy** : Presence of lymph node involvement (Yes or No).

-   **Pathology** : Type of thyroid cancer (e.g., Micropapillary).

-   **Focality** : Tumor focality (Uni-Focal or Multi-Focal).

-   **Risk** : Cancer risk classification (Low, Intermediate, High).

-   **T** : Tumor classification (T1, T2, etc.).

-   **N** : Lymph node classification (N0, N1, etc.).

-   **M** : Metastasis classification (M0, M1, etc.).

-   **Stage** : Cancer staging (Stage I, II, III, IV).

-   **Response** : Treatment response (Excellent, Indeterminate, etc.).

-   **Recurred** : Whether cancer recurred (Yes or No).

```{r, echo=FALSE}
library(glue)
library(tidymodels)
library(tidyverse)
theme_set(theme_classic())

setwd("C:/Users/Magdalena/portfolio")
thyroid_data <- read.csv("ph_data/thyroid_data.csv") 
thyroid_data <- thyroid_data |>
  mutate(across(where(is.character), as.factor))
str(thyroid_data)

```

What is the average age of a patient with thyroid cancer?

```{r}
mean_age <- thyroid_data |> 
  summarize(mean_age = mean(Age)) |>
  unlist() 

min(thyroid_data$Age)
max(thyroid_data$Age)

ggplot(thyroid_data, aes(x = Age)) +
  geom_histogram(bins = 20, fill = "lightblue", color= "black") +
  scale_x_continuous(breaks = seq(10, 90, by = 10)) +
  labs(title = "Age Histogram", x = "Age", y = "Number of Patients") +
  geom_vline(xintercept = mean_age, color = "red", linetype = "dashed") +
  annotate("text", x = 50, y = 45,
           label = glue("Average Age {round(mean_age, 2)}"), 
           color = "red")

ggplot(thyroid_data, aes(x = Age)) +
  geom_histogram(bins = 20, fill = "lightblue", color= "black") +
  scale_x_continuous(breaks = seq(10, 90, by = 10)) +
  labs(title = "Age Histogram", x = "Age", y = "Number of Patients") +
  geom_vline(xintercept = mean_age, color = "red", linetype = "dashed") +
  annotate("text", x = 50, y = 45,
           label = glue("Average Age {round(mean_age, 2)}"), 
           color = "red")
```

What are the proportion of female vs male patients?

```{r}
gender_counts <- thyroid_data |>
  count(Gender, name = "total")

total_pt <- nrow(thyroid_data)

glue("The percentage of female patients: {round(gender_counts[1,2]/total_pt * 100, 2)}%\n The percent of male patients: {round(gender_counts[2,2]/total_pt * 100, 2)}%")
```

What are the recurrence rates based on gender?

```{r}
recurred_sum <- thyroid_data |>
  summarize(recurred_sum = sum(as.numeric(Recurred)-1), .by = Gender)

recurred_graph <- right_join(recurred_sum, gender_counts, by = "Gender") |>
  mutate(no_recurred = total - recurred_sum)

recurred_graph1 <- recurred_graph |>
  pivot_longer(cols = c(recurred_sum, no_recurred),
               names_to = "Status", values_to = "Count")

ggplot(recurred_graph1, aes(x = Gender, y = Count, fill = Status)) +
  geom_col() +
  scale_fill_manual(
    values = c("recurred_sum" = "red3", "no_recurred" = "lightblue3"),
    labels = c("Recurred", "Did Not Recur"))
```

## Logistic regression model

```{r}
thyroid_data |>
  count(Recurred)

#split data 
set.seed(2)
split <- initial_split(thyroid_data, prop = 0.8, strata = Recurred)
train <- training(split)
test <- testing(split)

ggplot(train, aes(x = Age)) +
  geom_boxplot(aes(y = "Train")) +
  geom_boxplot(data = test, aes(y = "Test"))

vfold <- vfold_cv(train, strata = Recurred)

```

The data is split into 80% for training and 20% for testing. Since recurrence happens in about 40% of cases, the split is stratified on the Recurrence feature to ensure the test data contains enough cases of recurrence. The random split looks balanced between the test and training set when checked on the Age feature.

```{r}
recipe_recurred <- recipe(Recurred ~., data = thyroid_data) |>
  step_normalize(Age) |>
  step_dummy(all_nominal_predictors())
  
logistic_model <- logistic_reg() |>
  set_engine("glm") |>
  set_mode("classification") 

logistic_wf <- workflow() |>
  add_model(logistic_model) |>
  add_recipe(recipe_recurred)

logistic_fit <- fit(logistic_wf, data = train)
tidy(logistic_fit) |>
  filter(p.value <= 0.05)

logistic_predictions <- predict(logistic_fit, new_data = test)
logistic_matrix <- test |>
  select(Recurred) |>
  cbind(logistic_predictions) |>
  count(Recurred, .pred_class) |>
  rename(Predicted = .pred_class) |>
  print()
```

Using the logistic regression model to predict the recurrence of thyroid cancer based on all of the features available in the data we obtain 73 correct predictions, 3 false negatives, and 0 false positives.

## Random Forest Model

```{r}
rf_model <- rand_forest() |>
  set_mode("classification") |>
  set_engine("ranger")

rf_workflow <- workflow() |>
  add_recipe(recipe_recurred) |>
  add_model(rf_model)

rf_fit <- fit(rf_workflow, data = train)
rf_predictions <- predict(rf_fit, new_data = test)

test |>
  select(Recurred) |>
  cbind(rf_predictions) |>
  count(Recurred, .pred_class) |>
  rename(Predicted = .pred_class) |>
  print()
```

## XGBoost Model

```{r}
xgb_model <- boost_tree() |>
  set_mode("classification") |>
  set_engine("xgboost")

xgb_workflow <- workflow() |>
  add_model(xgb_model) |>
  add_recipe(recipe_recurred)

xgb_fit <- fit(xgb_workflow, data = train)
xgb_predictions <- predict(xgb_fit, new_data = test)

test |>
  select(Recurred) |>
  cbind(xgb_predictions) |>
  count(Recurred, .pred_class) |>
  rename(Predicted = .pred_class) |>
  print()
```

## Validation

```{r}
workflow_names <- c("logit", "rf", "xgboost")
workflow_variable <- list(logistic_wf, rf_workflow, xgb_workflow)
class_metric <- metric_set(accuracy, brier_class, roc_auc)

workflows <- tibble("workflow_name" = workflow_names,
                    "workflow_object" = workflow_variable) |>
  rowwise()|>
  mutate(fits = list(fit_resamples(workflow_object,
                                   vfold,
                                   metrics = class_metric))) |>
  mutate(metrics = list(collect_metrics(fits)))
```

## Model Assessment

```{r}

```
